---
title: 'Getting Started With R Package Development'
author: "Fernando Abarca"
date: '2017-05-22'
output:
  html_document: default
  pdf_document: default
---

-------------------------------------------------------------------------------

Tutorial Source Code: [github.com/fabarca/r-package-tutorial](https://github.com/fabarca/r-package-tutorial)

Main Reference: [R Packages - by Hadley Wickham](http://r-pkgs.had.co.nz/)

```{r functions, include=FALSE}
print.files <- function(path = "."){
  cat(
      sort(
          c(paste0(list.dirs(path, full.names=F)[-1], "/"),
            list.files(path, full.names=F, recursive=T)
            )
        ), 
      sep='\n'
    )
}
```


----

## Index {#index}

0. [Basic Commands](#commands)
1. [Introduction](#intro)
2. [Requirements](#requirements)
3. [Package Creation](#creation)
4. [Adding Functions](#functions)
5. [Documentation](#documentation)
6. [Installation](#installation)
7. [Usage Example](#using)
8. [Testing](#tests)
9. [Vignettes](#vignettes)


----

## Basic Commands {#commands}
[Return to index &uarr;](#index)

This is a cheat sheet for creating packages in R. These commands will be explained along this tutorial.

```{r basic_commands, eval=FALSE}
# The first argument of the following functions must be the path to the package
# but if you are already in the right folder, it isn't necessary to specify it.
# To be sure, you may set the working directory by using this command:
setwd("path/packageName")

# Create package structure
devtools::create(".")
# Add some functions inside R/ folder, 
# then create documentation and NAMESPACE file
devtools::document()
# Now you can install your packages
devtools::install()

# Create test folder
devtools::use_testthat()
# Add some tests inside tests/testthat/,
# then perform your tests
devtools::test()

# Add a vignette template and dependencies
devtools::use_vignette("vignette-name")
# Install including vignettes
devtools::install(build_vignettes = TRUE)
```

----

## Introduction {#introduction}
[Return to index &uarr;](#index)

The aim of this tutorial is to provide a quick start guide to R users interested in creating packages by learning only the essentials. 
The user is expected to already know how to code in R.

For more details about package creation, please refer to the main reference link.

This Tutorial was written using Rmarkdown, and it actually creates a R package while is built! Both, the Rmd file and the R package are available in [github.com/fabarca/r-package-tutorial](https://github.com/fabarca/r-package-tutorial).

----

## Requirements {#requirements}
[Return to index &uarr;](#index)

These are the packages that you need to install to get started:
```{r required_packages, eval=FALSE}
install.packages(c("devtools", "roxygen2", "testthat", "knitr"))
```

To have the last version of devtools you may use:
```{r install_devtools, eval=FALSE}
devtools::install_github("hadley/devtools")
```

```{r clean_dir, include=FALSE}
# Clean example directory
unlink("MyRPackages/firstPkg", recursive = TRUE)
if(!dir.exists("./MyRPackages")){
  dir.create("./MyRPackages")
}
```

-------------------------------------------------------------------------------

## Package Creation {#creation}
[Return to index &uarr;](#index)

First, create a new package named `firstPkg` inside `MyRPackages/` folder by using:
```{r create_package, message=FALSE, warning=FALSE}
devtools::create("MyRPackages/firstPkg")
```

Among the files created, will be the `DESCRIPTION` file, that you can edit with your own information.

-------------------------------------------------------------------------------

## Adding Functions {#functions}
[Return to index &uarr;](#index)

This is the basic R package structure:

```{r list_dir, echo=FALSE}
print.files("MyRPackages/firstPkg")
```

The folder `R/` is empty at this moment, all your functions should be placed there.  
For example, you can create a script file called `groupByDate.R` with following code inside:

```{r, code=readLines("examples/groupByDate.R"), eval=FALSE}
```

```{r copy_script_r, include=FALSE}
file.copy("examples/groupByDate.R", "MyRPackages/firstPkg/R/groupByDate.R")
```


----

## Documentation {#documentation}
[Return to index &uarr;](#index)

All comments starting by `#'` are the metadata used by roxygen2 package to automatically create the documentation files (inside the folder `man/`) and fill the NAMESPACE file.  
Before the function definition, you need to add `#' @export` , in order to make the function available in the `NAMESPACE` file. This will make the function publicly available, otherwise it would be only accesible from functions within the package.  
In this example, you will need to use some functions from a external package. To do that, add a comment like this `#' @importFrom other_package function1 function2`.


To create documentation based on roxygen2 comments, run the following command:
```{r hidden_create_documentation, include=FALSE}
devtools::document("MyRPackages/firstPkg")
```

```{r create_documentation, eval=FALSE}
devtools::document("MyRPackages/firstPkg")

## Updating firstPkg documentation
## Loading firstPkg
## Updating roxygen version in MyRPackages/firstPkg/DESCRIPTION
## Writing NAMESPACE
## Writing groupByDate.Rd
```

Now your package folder should look like this:
```{r list_files_doc, echo=FALSE}
print.files("MyRPackages/firstPkg")
```


Remember to add the external libraries that you need to import in the `DESCRIPTION` file. 
In this case you should add this line: `Imports: dplyr`
```{r, include=FALSE}
file.copy("examples/DESCRIPTION", "MyRPackages/firstPkg/DESCRIPTION", overwrite = TRUE)
```


-------------------------------------------------------------------------------

## Installation {#installation}
[Return to index &uarr;](#index)

To install this package run:
```{r hidden_install, include=FALSE}
devtools::install("MyRPackages/firstPkg")
```
```{r install, eval=FALSE}
devtools::install("MyRPackages/firstPkg")
```

Now you can check the documentation page of your function using:
```{r help_function, eval=FALSE}
?groupByDate
```

----

## Usage Example {#using}
[Return to index &uarr;](#index)

Here is an example code to try the new function:
```{r using_function}
library(firstPkg)
set.seed(1)

# Create a dataframe
date_vec = rep(as.Date('2017-01-01') + 0:8, 4)
value_vec = sample(1:36 * 4, 36)
example_df = data.frame(date = date_vec, value = value_vec)

# Group the dataframe using the new function
firstPkg::groupByDate(example_df)
```

----

## Testing {#tests}
[Return to index &uarr;](#index)

To create a test folder using testthat package run:

```{r create_testthat, message=FALSE, warning=FALSE}
devtools::use_testthat("MyRPackages/firstPkg")
```

Here you can see the new folders:

```{r r_list_test_dir, echo=FALSE}
print.files("MyRPackages/firstPkg/")
```

Tests must go inside `tests/testthat/` folder with the prefix `test`. For example, you can create a new test script called `test_groupByDate.R`, with the following code inside:

```{r, code=readLines("examples/test_groupByDate.R"), eval=FALSE}
```

```{r, include=FALSE}
file.copy("examples/test_groupByDate.R", "MyRPackages/firstPkg/tests/testthat/test_groupByDate.R")
```

To perform the tests run:
```{r run_tests, message=FALSE, warning=FALSE}
devtools::test("MyRPackages/firstPkg")
```

This is the current directory structure, where almost everything was created automatically. You only had to manually add the `R/groupByDate.R` and `tests/testthat/test_groupByDate.R` files, and edit the `DESCRIPTION` file.
```{r list_files_final, echo=FALSE}
print.files("MyRPackages/firstPkg")
```

----

## Vignettes {#vignettes}
[Return to index &uarr;](#index)

You may also wish to include a vignette in your package. For example to create a new vignette called `firstPkg_intro` run:

```{r code_create_vignette, eval=FALSE}
devtools::use_vignette("firstPkg_intro", "MyRPackages/firstPkg")
```

```{r run_create_vignette, include=FALSE}
devtools::use_vignette("firstPkg_intro", "MyRPackages/firstPkg")
```

This command creates a template file called `firstPkg_intro.Rmd` inside `vignettes/` folder. It also automatically adds the dependencies in the `DESCRIPTION` file.

```{r r_list_vignettes_dir, echo=FALSE}
print.files("MyRPackages/firstPkg")
```

Now you can edit this template with your own code, for example:
```{r, echo=FALSE, comment=''}
cat(readLines('examples/firstPkg_intro.Rmd'), sep = "\n")
```

```{r copy_vignette, include=FALSE}
file.copy("examples/firstPkg_intro.Rmd", "MyRPackages/firstPkg/vignettes/firstPkg_intro.Rmd", overwrite = TRUE)
```

To install the package including vignettes run:
```{r hidden_install_vignettes, include=FALSE}
devtools::install("MyRPackages/firstPkg", build_vignettes = TRUE)
```
```{r install_vignettes, eval=FALSE}
devtools::install("MyRPackages/firstPkg", build_vignettes = TRUE)
```

To open the vignette use:
```{r, eval=FALSE}
vignette('firstPkg_intro')
```

```{r adding_readme, include=FALSE}
cat("This package was automatically created while the tutorial compiling.\n", file = './MyRPackages/firstPkg/README.md')
```

----


#### Additional references:  

[rmarkdown-cheatsheet (PDF)](https://www.rstudio.com/wp-content/uploads/2016/03/rmarkdown-cheatsheet-2.0.pdf) 

[rmarkdown-reference (PDF)](https://www.rstudio.com/wp-content/uploads/2015/03/rmarkdown-reference.pdf)




